Clustering time series based on probability distributions across temporal granularities

### 2021-11-02 main.Rmd
proof reading changes

abstract: |
 With more and more time series data being collected at much finer temporal resolution, for a longer length of time, and for a larger number of individuals/entities, time series clustering research is getting a lot of traction. Long, noisy, patchy, uneven, and asynchronous time series are common in many fields, limiting similarity searches or lowering method efficiency when clustering is based on a distance metric. In this work, we suggest two approaches for obtaining similarity between time series based on probability distributions over cyclic temporal granularities for distance-based clustering approaches. Cyclic granularities like hour-of-the-day, work-day/weekend, month-of-the-year and so on, are useful for finding repeated patterns in the data. Looking at probability distributions across cyclic granularities serves two purposes: (a) "Probability distributions" characterise the inherent temporal data structure of these large unequal-length time series and are robust to missing or noisy data. (b) Using probability distributions over "cyclic granularities" ensures small pockets of similar "repeated" behaviours. Our method is capable of producing useful clusters, as demonstrated by testing on validation data designs and a sample of residential smart metee consumers.

@@

```{r mytheme}
theme_validation <- function() {
  theme_bw() +
      theme(
    strip.text = element_text(size = 8, margin = margin(b = 0, t = 1)),
    plot.margin = margin(0, 0, 0, 0, "cm"),
    axis.title.y = element_blank(),
    #axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.ticks = element_blank(),
      legend.position = "bottom"
    )
}

theme_characterisation <- function() {

  theme_bw() + # seeting theme
    theme(strip.text = element_text(size = 10,
                                    margin = margin(b = 0, t = 0))) + # narrow facet space
   theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + # no axis ticks 
  theme(panel.spacing =unit(0, "lines")) +  # to ensure no gap between facets
    theme(axis.text.x = element_text(angle=90, hjust=1, size = 10)) + # rotate the x-axis text
    theme(legend.position = "bottom")+
  theme(plot.margin = margin(0, 0, 0, 0, "cm")) +
  theme(axis.text.x = element_text(size=5))
}
```

```{r mytheme-application}
theme_application <- function() {

  theme_light() + # setting theme
    #theme(strip.text = element_text(margin = margin(b = 0, t = 0))) + # narrow facet space
   theme(axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + # no axis ticks 
  theme(panel.spacing =unit(0, "lines")) +  # to ensure no gap between facets
    theme(axis.ticks.x = element_blank(),
        axis.text.x = element_blank()) + # no x-axis labels to further reduce the gap between facets
    #theme(axis.text.x = element_text(angle=90, hjust=1, size = 9)) + # rotate the x-axis text
  theme(plot.margin = margin(0, 0, 0, 0, "cm")) +
  #theme(axis.text.x = element_text(size=5)) +
    theme(strip.background = element_blank(),
  strip.text.x = element_blank())
}
```

@@

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) meaningful groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance strategy to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularization, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

\noindent We consider the problem of clustering a large number of univariate time series of continuous values which are available at fine temporal scales, being motivated by the residential smart meter data. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and missing observations and characterized by noisy and patchy behavior that can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Choosing probability distributions seem to be a natural way to analyze these types of data sets since they are robust to uneven length, missing data, or noise. Hence this paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across significant cyclic granularities. Cyclic temporal granularities, which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across each of the interesting cyclic granularities. Below are some of the benefits of our method, which will be detailed in further depth in subsequent sections.

- When using probability distributions, data does not have to be the same length or observed during the exact same time period (unless there is a structural pattern).

@@

A multitude of papers have emerged around smart meter time series clustering for deepening our knowledge of consumption patterns. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilized in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency . @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time-series common in residential  and addresses model-based clustering  by transforming the series into other other objects such as structure or set of parameters which can be more easily characterized and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. None of the methods focus on exploring heterogeneity in repetitive behaviors across different temporal dependencies, we present a way to do that in this paper.

@@

- Remove customers whose data for an entire category of a significant granularity is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.

@@

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature, yet it could be a useful tool for evaluating model or method performance. There are several ways to approach the prototype selection.
Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers who are far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that was hidden in a single variable display better than static projections. Hence we perform a linked tour with t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. Please see the Supplementary article for further details on how the prototypes are chosen. Figure \ref{fig:hod-ind-group} (a, b, c) shows the raw time plot, distribution across `hod`, `moy` and `wkndwday` for the set of chosen $24$ customers. Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution.

@@

wkndwday_ind_design <- data_wkndwday%>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(design), 
   #                   color = as.factor(design)), k=5, alpha = 0.5) +
  geom_boxplot(alpha = 0.5, fill = "black")+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
              nrow = 24)  +
  stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 1, color = "black")+
  xlab("wdwn")+
  theme_application()

@@

Cluster characterization is a crucial aspect of cluster analysis. The $24$ prototypes <!--and the data subset consisting of $353$ customers--> are clustered using the methodology described in \ref{sec:methodology} and results are reported below. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$. All customers with the same color represent same clustered groups. Groups by JS-based distances and wpd-based distances are colored differently as they represent different groupings. The plotting scales are not displayed since we want to emphasize comparable shapes rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to shape or significance across cyclic granularities.

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The distribution of electricity demand for the selected $24$ customers across hod, moy and wdwn are shown in Figure \ref{fig:hod-ind-group} (d, e, f). Our methodology is useful for grouping similar distributions over hod and moy and they are placed closely for easy comparison. Of course, certain customers in each group have distributions that differ from other members in the same group. However, it appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. Figure \ref{fig:combined-groups-js} shows Groups $2$ and $5$ show a stronger hod pattern with a typical morning and evening peak, whereas groups 1, 3, and 5 show a moy pattern  with higher usage in winter months. Differences in Weekend/Weekday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters.

@@

wkndwday_ind_group <- data_wkndwday%>% 
  left_join(cluster_result, by = c("customer_id")) %>% 
  mutate(group  = as.factor(group)) %>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(group), 
          #            color = as.factor(group)), k=5, alpha = 0.5) +
  geom_boxplot(aes(color = group, fill = group),alpha = 0.5)+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(group~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
             nrow=24)  +
  theme_application()+
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7"))  + theme(legend.position = "none")+
  xlab("wdwn")

@@

We chose the optimal number of clusters using (@Hennig2014-ah) as $3$. A parallel coordinate plot with the three significant cyclic granularities used for wpd-based clustering. The variables are sorted according to their separation across classes (rather than their overall variation between classes). This means that $moy$ is the most important variable in distinguishing the designs followed by $hod$ and $wkndwday$. There is only one customer who has significant $wpd$ across $wkndwday$ and stands out from the rest of the customers. Group 3 has a higher wpd for hod than moy or wkndwday. Group 2 has the most distinct pattern across moy. Group 1 is a mixed group that has strong patterns on at least one of the three variables. The findings vary from js-based clustering, yet it is a helpful grouping.

@@

data_table <- data_pcp %>% group_by(group) %>% 
  summarise(nobs = n(),
            moy = round(median(month_year),2),
            hod = round(median(hour_day),2),
            wdwn = round(median(wknd_wday),2)) %>% 
  select(-group)

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining heterogeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the Weekend/Weekday behavior in the given case does not characterize any cluster. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our clustering methodology to cluster them, and demonstrate that the method produces useful clusters.

@@

We offer two methods for calculating pairwise distances between time series based on probability distributions at various cyclic granularities. Depending on the goal of the clustering, these distance metrics, when fed into a hierarchical clustering algorithm using Ward's linkage, yield meaningful clusters. Probability distributions provide an intuitive method to characterise noisy, patchy, long, and unequal-length time series data. Distributions over cyclic granularities help to characterise the formed clusters in terms of their repeating behavior over these cyclic granularities. Furthermore, unlike earlier efforts that group customers based on behavior across only one cyclic granularity (such as hour-of-day), our method is more comprehensive in detecting clusters with repeated patterns at all relevant granularities. The JS-based approaches are faster than wpd-based approaches.

There are few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, we have assumed the time series to be stationary, and hence the distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach can not detect these dynamic changes. Thirdly, it is possible that for few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers in the clustering by handling their missing categories. Finally, $wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers.

>>>>

abstract: |
 With more and more time series data being collected at much finer temporal resolution, for a longer length of time, and for a larger number of individuals/entities, time series clustering research is getting a lot of traction. Long, noisy, patchy, uneven, and asynchronous time series are common in many fields, limiting similarity searches or lowering method efficiency when clustering is based on a distance metric. In this work, we suggest two approaches for obtaining similarity between time series based on probability distributions over cyclic temporal granularities for distance-based clustering approaches. Cyclic granularities like hour-of-the-day, work-day/weekend, month-of-the-year and so on, are useful for finding repeated patterns in the data. Looking at probability distributions across cyclic granularities serves two purposes: (a) "Probability distributions" characterise the inherent temporal data structure of these large unequal-length time series and are robust to missing or noisy data. (b) Using probability distributions over "cyclic granularities" ensures small pockets of similar "repeated" behaviours. Our method is capable of producing useful clusters, as demonstrated by testing on validation data designs and a sample of residential smart meter consumers.

@@

```{r mytheme-application}
```

```{r theme_characterisation}
```

```{r theme-validation}

@@

<!-- time series clustering and its challenges -->
Time-series clustering is the process of unsupervised partitioning of $n$ time-series data into $k$ ($k<n$) meaningful groups such that homogeneous time-series are grouped together based on a certain similarity measure. The time-series features, length of time-series, representation technique, and, of course, the purpose of clustering time-series all influence the suitable similarity measure or distance metric to a meaningful level. The three primary methods to time series clustering (@liao2005clustering) are algorithms that operate directly with distances or raw data points in the time or frequency domain (distance-based), with features derived from raw data (feature-based), or indirectly with models constructed from raw data (model-based). The efficacy of distance-based techniques is highly dependent on the distance measure utilized. Defining an appropriate distance measure for the raw time series may be a difficult task since it must take into account noise, variable lengths of time series, asynchronous time series, different scales, and missing data. Commonly used Distance-based similarity measures as suggested by a decade review of time series clustering approaches (@Aghabozorgi2015-ct) are Euclidean, Pearson's correlation coefficient and related distances, Dynamic Time Warping, Autocorrelation, Short time series distance, Piecewise regularization, cross-correlation between time series, or a symmetric version of the Kullback–Liebler distances (@liao2007clustering) but on a vector time series data. Among these alternatives, Euclidean distances have high performance but need the same length of data over the same period, resulting in information loss regardless of whether it is on raw data or a smaller collection of features. DTW works well with time series of different lengths (@corradini2001dynamic), but it is incapable of handling missing observations. Surprisingly, probability distributions, which may reflect the inherent temporal structure of a time series have not been considered in determining time series similarity.

\noindent We consider the problem of clustering a large number of univariate time series of continuous values which are available at fine temporal scales, being motivated by the residential smart meter data. These time series data are long (with more and more data collected at finer resolutions), are asynchronous, with varying time lengths for different houses and missing observations and characterized by noisy and patchy behavior that can quickly become overwhelming and hard to interpret, requiring summarizing the large number of customers into pockets of similar energy behavior. Choosing probability distributions seem to be a natural way to analyze these types of data sets since they are robust to uneven length, missing data, or noise. This paper proposes two approaches for obtaining pairwise similarities based on Jensen-Shannon distances between probability distributions across significant cyclic granularities. Cyclic temporal granularities [@Gupta2021-hakear], which are temporal deconstructions of a time period into units such as hour-of-the-day, work-day/weekend, can be useful for measuring repetitive patterns in large univariate time series data. The resulting clusters are expected to group customers that have similar repetitive behaviors across each of the interesting cyclic granularities. Below are some of the benefits of our method:

- When using probability distributions, data does not have to be the same length or observed during the exact same time period (unless there is a structural pattern);

@@

A multitude of papers have emerged around smart meter time series clustering for deepening our knowledge of consumption patterns. @Tureczek2017-pb conducted a systematic study of over $2100$ peer-reviewed papers on smart meter data analytics. None of the $34$ articles chosen for their emphasis use Australian smart meter data. The most often used algorithm is K-Means. Using K-Means without considering time series structure or correlation results in inefficient clusters. Principal Component Analysis (PCA) or Self-Organizing Maps (SOM) eliminate correlation patterns and decrease feature space, but lose interpretability. To reduce dimensionality, several studies use principal component analysis or factor analysis to pre-process smart-meter data before clustering (@Ndiaye2011-pf). Other algorithms utilized in the literature include k-means variants, hierarchical approaches, and greedy k-medoids. Time series data, such as smart meter data, are not well-suited to any of the techniques mentioned in @Tureczek2017-pb. Only one study [@ozawa2016determining] identified time series characteristics using Fourier transformation, which converts data from time to frequency and then uses K-Means to cluster by greatest frequency. @Motlagh2019-yj suggests that the time feature extraction is limited by the type of noisy, patchy, and unequal time-series common in residential  and addresses model-based clustering  by transforming the series into other other objects such as structure or set of parameters which can be more easily characterized and clustered. [@chicco2010renyi] addresses information theory-based clustering such as Shannon or Renyi entropy and its variations. @Melnykov2013-sp discusses how outliers, noisy observations and scattered observations can complicate estimating mixture model parameters and hence the partitions. In our knowledge, none of the methods focus on exploring heterogeneity in repetitive behaviors based on the dynamics of multiple temporal dependencies using probability distributions.

@@

- Remove customers whose data for an entire category of `hod`, `moy` or `wnwd` is empty. For example, a customer who does not have data for an entire month is excluded because their monthly behavior cannot be analyzed.

@@

<!-- Why instance selection -->
Supervised learning uses a training set of known information to categorize new events through instance selection. Instance selection (@olvera2010review) is a method of rejecting instances that are not helpful for classification. This is analogous to subsampling the population along all dimensions of interest such that the sampled data represents the primary features of the underlying distribution. Instance selection in unsupervised learning has received little attention in the literature, yet it could be a useful tool for evaluating model or method performance. There are several ways to approach the prototype selection.
Following @Fan2021-bq's idea of picking related examples (neighbors) for each instance (anchor), we can first use any dimensionality reduction techniques like MDS or PCA to project the data into a 2D space. Then pick a few "anchor" customers who are far apart in 2D space and pick a few neighbors for each. Unfortunately, this does not ensure that consumers with significant patterns across all variables are chosen. Tours can reveal variable separation that was hidden in a single variable display better than static projections. Hence we perform a linked tour with t-SNE layout using the R package `liminal` (@R-liminal) to identify customers who are more likely to have distinct patterns across the variables studied. (Refer to Supplementary article for further details). Figure \ref{fig:hod-ind-group} (a, b, c) shows the raw time plot, distribution across `hod`, `moy` and `wnwd` for the set of chosen $24$ customers. Few of these customers have similar distribution across `moy` and some are similar in their `hod` distribution.

@@

wkndwday_ind_design <- data_wkndwday%>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(design), 
   #                   color = as.factor(design)), k=5, alpha = 0.5) +
  geom_boxplot(alpha = 0.5, fill = "black")+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
              nrow = 24)  +
  stat_summary(
    fun = median,
    geom = 'line',
    aes(group = 1), size = 1, color = "black")+
  xlab("wnwd")+
  theme_application()

@@

The $24$ prototypes <!--and the data subset consisting of $353$ customers--> are clustered using the methodology described in Section \ref{sec:methodology} and results are reported below. In the following plots, the median is shown by a line, and the shaded region shows the area between the  $25^{th}$ and $75^{th}$. All customers with the same color represent same clustered groups. Groups by JS-based distances and wpd-based distances are colored differently as they represent different groupings. The plotting scales are not displayed since we want to emphasize comparable shapes rather than scales. The idea is that a customer in a cluster may have low total energy usage, but their behavior may be quite similar to a customer with high usage with respect to distributional pattern or significance across cyclic granularities.

@@

For clustering based on JS-based distances, we chose the optimal number of clusters using (@Hennig2014-ah) as $5$. The distribution of electricity demand for the selected $24$ customers across `hod`, `moy` and `wnwd` are shown in Figure \ref{fig:hod-ind-group} (d, e, f). Our methodology is useful for grouping similar distributions over hod and moy and they are placed closely for easy comparison. Of course, certain customers in each group have distributions that differ from other members in the same group. However, it appears that the aim of grouping comparable distributions over considered variables has been accomplished to some extent. Figure \ref{fig:combined-groups-js} shows the summarized distributions across $5$ groups and assists us in characterizing each cluster. It shows Groups $2$ and $5$ show a stronger hod pattern with a typical morning and evening peak, whereas groups 1, 3, and 5 show a moy pattern  with higher usage in winter months. Differences in Weekend/Weekday between groups are not discernible, implying that it may not be a relevant variable in distinguishing various clusters unless maybe conditioned by `moy` or `hod`.

@@

wkndwday_ind_group <- data_wkndwday%>% 
  left_join(cluster_result, by = c("customer_id")) %>% 
  mutate(group  = as.factor(group)) %>% 
  ggplot(aes(x=wknd_wday, y = general_supply_kwh)) +
  #lvplot::geom_lv(aes(fill = as.factor(group), 
          #            color = as.factor(group)), k=5, alpha = 0.5) +
  geom_boxplot(aes(color = group, fill = group),alpha = 0.5)+
  #geom_boxplot(outlier.shape = NA) + 
  coord_cartesian(ylim = ylim1*1.05)+
  facet_wrap(group~customer_id, 
             scales = "free_y", 
             labeller = "label_value",
             nrow=24)  +
  theme_application()+
  scale_fill_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00","#CC79A7"))+
  scale_color_manual(values = c("#E69F00", "#009E73","#0072B2", "#D55E00", "#CC79A7"))  + theme(legend.position = "none")+
  xlab("wnwd")

@@

We chose the optimal number of clusters using (@Hennig2014-ah) as $3$. A parallel coordinate plot with the three significant cyclic granularities is used to characterise the groups here. The variables are sorted according to their separation across classes (rather than their overall variation between classes). This means that $moy$ is the most important variable in distinguishing the designs followed by $hod$ and $wnwd$. There is only one customer who has significant $wpd$ across $wnwd$ and stands out from the rest of the customers. Group 3 has a higher wpd for hod than moy or wkndwday. Group 2 has the most distinct pattern across moy. Group 1 is a mixed group that has strong patterns on at least one of the three variables. The findings vary from js-based clustering, yet it is a helpful grouping.

@@

data_table <- data_pcp %>% group_by(group) %>% 
  summarise(nobs = n(),
            moy = round(median(month_year),2),
            hod = round(median(hour_day),2),
            wnwd = round(median(wknd_wday),2)) %>% 
  select(-group)

@@

Things become far more complicated when we consider a larger data set with more uncertainty, as they do with any clustering problem. Summarizing distributions across clusters with varied or outlying customers can result in a shape that does not represent the group. Furthermore, combining heterogeneous customers may result in similar-looking final clusters that are not effective for visually differentiating them. It is also worth noting that the Weekend/Weekday behavior in the given case does not characterize any cluster. This, however, will not be true for all of the customers in the data set. If more extensive prototype selection is used, resulting in more comprehensive prototypes in the data set, this method might be used to classify the entire data set into these prototype behaviors. However, the goal of this section was to have a few customers that have significant patterns over one or more cyclic granularities, apply our methodology to cluster them, and demonstrate that the method produces useful clusters.

@@

We offer two approaches for calculating pairwise distances between time series based on probability distributions over multiple cyclic granularities at once. Depending on the goal of the clustering, these distance metrics, when fed into a hierarchical clustering algorithm using Ward's linkage, yield meaningful clusters. Probability distributions provide an intuitive method to characterise noisy, patchy, long, and unequal-length time series data. Distributions over cyclic granularities help to characterise the formed clusters in terms of their repeating behavior over these cyclic granularities. Furthermore, unlike earlier efforts that group customers based on behavior across only one cyclic granularity (such as hour-of-day), our method is more comprehensive in detecting clusters with repeated patterns at all relevant granularities. <!--The JS-based approaches are faster than wpd-based approaches.-->

There are few areas to extend this research. First, larger data sets with more uncertainty complicate matters, as is true for any clustering task. Characterizing clusters with varied or outlying customers can result in a shape that does not represent the group. Moreover, integrating heterogeneous consumers may result in visually identical end clusters, which are potentially not useful. Hence, a way of appropriately scaling it up to many customers such that anomalies are removed before clustering would be useful for bringing forth meaningful, compact and separated clusters. Secondly, we have assumed the time series to be stationary, and hence the distributions are assumed to remain constant for the observation period. In reality, however, it might change. For the smart meter example, the distribution for a customer moving to a different house or changing electrical equipment can change drastically. Our current approach can not detect these dynamic changes. Thirdly, it is possible that for few customers, data for some categories from the list of considered significant granularities are missing. In our application, we have removed those customers and done the analysis but the metrics used should be able to incorporate those customers in the clustering by handling their missing categories. Finally, $wpd$ is computationally heavy even under parallel computation. Future work can make the computations more efficient so that they are easily scalable to a large number of customers. Moreoever, experiments can also be run with non-hierarchy based clustering algorithms to verify if these distances work better with other algorithms.
